FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-06-07 paper.tex
finch case study, slicing future work

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{amsmath}

@@

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FlowFPX}{FlowFPX}

@@

To illustrate what FloatTracker is capable of, we applied our tool to three different Julia libraries to help us track down strange behavior:

\begin{description}
\item[Surprise NaNs] We found that otherwise innocuous values for a time step parameter in the \texttt{ShallowWaters} library can produce NaNs in the result.
  We used FloatTracker to track down where the NaNs arose.
\item[NaN Fuzzing] We injected NaNs in a N-Body simulation and were able to find a bug in the widely-used \texttt{OrdinaryDiffEq} library.
\item[NaN Kills] We observed issues with NaN kills inside of the \texttt{Finch} library.
  % FIXME: more needed here
\end{description}

All three libraries are available from the JuliaHub package archive.
The \texttt{OrdinaryDiffEq} library in particular enjoys widespread use in the scientific computing community.

@@

\subsection{NaN Kills: Finch}

% Advection examples go here, right?

@@

>>>>

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{amsmath}
\usepackage{cleveref}

@@

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FlowFPX}{FlowFPX}
\newcommand{\FloatTracker}{FloatTracker}
\newcommand{\FT}{\FloatTracker}
\newcommand{\CSTG}{stack graph}
\newcommand{\CPP}{\code{C++}}
\newcommand{\Dendro}{\textsc{Dendro}}
\newcommand{\urlaccess}[2]{\url{#1}}

@@

\FT{} is an effective tool for debugging all sorts of NaN-related issues.
We have used it to reveal silent NaNs, uncover the root cause of NaNs,
and fuzz for weak points in several case studies.

@@

\subsection{Finch}
\label{s:finch}

Finch is a domain-specific language for specifying
PDEs~\cite{heislerFinchDomainSpecific2022}.\footnote{Not to be confused
with the Finch loop optimizer~\cite{adka-cgo-2023}.}
In the spirit of FEniCS~\cite{fenics} and related
tools~\cite{freefem,openfoam,dune,firedrake},
Finch helps scientists quickly convert math into code.
What sets Finch apart is its flexibility.
It supports multiple discretization methods (finite element and finite
volume) and multiple backends (Julia, \CPP{}, \Dendro{}~\cite{dendro}).
Furthermore it strives to output code that humans can easily fine-tune.

Fuzzing with \FT{} revealed two places where Finch needed
protection against user input.
The first was when reading an input mesh.\footnote{\urlaccess{https://github.com/paralab/Finch/issues/16}{2023-06-06}}
A NaN injected in the mesh led to a crash further on:
\begin{verbatim}
  BoundsError: attempt to access 1-element
   Vector{Int64} at index [2]
\end{verbatim}
Defensive programming solves the issue.
The second place was in setting bounds for the
solver.\footnote{\urlaccess{https://github.com/paralab/Finch/issues/17}{2023-06-06}}
Here, a NaN could leave bounds uninitialized, leading to a bounds error.
Additionally, \FT{} and \CSTG{}s have been useful for identifying NaNs that
appear in unstable systems.
%% advection2d, high cfl and high T (time) led to NaN during simulation

@@

\subsection{Enhanced Fuzzing}

While fuzzing is useful for discovering issues, its success rate
is low because \emph{every} floating-point operation is a candidate
for injection.
Even operations that are already well-defended against NaNs are candidates.
\FT{} could use two sorts of tools for improving injection.
First, fine-grained control to let users decide where not to inject.
Second, tools for understanding the context of an injection point
after the fact.
Program slicing is especially relevant to the latter point (and effectively
what we did by hand when fuzzing Finch~\cref{s:finch}).
For each operation, an expert needs to study the values that feed into it to
decide whether they are protected or not.

%% TODO example, start here at glbvertex https://github.com/paralab/Finch/blob/master/src/grid.jl#L360
